Hydras and Cie \\ Variations on Kirby \& Paris' hydra battles and ordinal numbers in Coq

### 2021-09-23 doc/JFLA2022/paper.tex
Minor copyediting

@@

Hydra games (also known as \emph{Hydra battles}) appear in an article published in 1982 by two mathematicians, 
Laurie Kirby and Jeff Paris: \emph{Accessible Independence Results for Peano Arithmetic}~\cite{KP82}.
This article describes a game between two players: Hercules and a hydra.
A short description of the game  can be found in~\cite{bauer2008, KP82, JFLA2018paper}. One can also play with
Andrej Bauer's simulator~\cite{BauerHydra}.
In a few words:
\begin{itemize}
\item A hydra is a finite tree, traditionally presented with the root at the bottom, the leaves of which are called \emph{heads}
  (Fig~\ref{fig:round}).
\item At every round, Hercules chops off one head of the hydra. If the head is at a distance greater or equal than 2 from the root,
  then some sub-tree $h$ of the hydra is copied a certain amount $n$ of times. The number $n$ of copies and the sub-tree $h$ may depend of the considered variant of the game
  or the time elapsed since the beginning of the fight.
  Figure~\ref{fig:round} shows an example with $n=2$.
\end{itemize}

@@

\begin{theorem}
  In the Hydra game, every strategy is a winning strategy for Hercules. \label{kp:thm1}
\end{theorem}

@@

Complex formalisations and proofs are explained in an
  electronic book~\cite{HydraBook}. Whenever various reasonable choices exist, we try to present and compare the alternatives.
  For instance, Figures~\ref{fig:Ex42E0} and \ref{fig:Ex42-schutte} show two radically different proofs of the equality
  $\omega+42+\omega^2=\omega^2$, the first one is a simple proof by computation, the second one shows how this equality
  is a consequence of the axioms of the set-theoretic model  by Kurt Schütte~\cite{schutte}.

@@

 Prior stages of this project have already been presented at
 JFLA~\cite{PCiota, JFLA2018paper}.
We present the recent evolutions of the library: new results, interaction within the Coq-Community project~\cite{CoqCommunity}, and documentation generated with Alectryon~\cite{alectryonpaper, alectryongithub}.

@@

Indeed, the consideration of \emph{all} battles allowed us to
consider battles where the hydra can make an arbitrary number of copies at any time, which made artificially easier our proof by absurd.
Unfortunately, the examples  most shown in the litterature
(see for instance \cite{KP82, bauer2008, BauerHydra}) 
assume that the hydra grows $n$ copies at step $n$ of the game, which is incompatible with our proof.

@@

We prove now that Theorem~\ref{thm3} still holds with this ``usual'' kind of battles, by borrowing new combinatorial results from~\cite{KS81}.
Without loss of generality, we could consider the following restrictions:

@@

We also study the number of steps of a battle:
Let $\alpha<\epsilon_0$ be an ordinal.
Using the \texttt{coq-equations} plug-in~\cite{sozeau:hal-01671777}, we define the function which returns the number of steps of the battle starting with $\alpha$ at step $i$, for  any $i$ in $\mathbb{N}$.
We prove that this number is greater or equal than
$H'_\alpha(i)-i$, where $H'$ is a slight variant of the Wainer Hardy hierarchy of rapidly growing functions~\cite{BW85, KS81, Promel2013, Wainer1970}, defined by transfinite recursion in Fig.~\ref{fig:Hprime}.
From this result, we infer that the considered function is not primitive recursive in general (for $\alpha\geq\omega^\omega$).

@@

@@

In order to prove formally that the length of the considered
kind of battles is not given by a primitive recursive function, weused the formalization of primitive recursive functions, part
of Russell O'Connor's contribution on G\"{o}del first incompleteness theorem \cite{OConnor05, Goedel}.
For this purpose, and above all by consideration of the scientific interest of this contribution, we chose to host and maintain this work within \textit{Coq-Community}.

@@

  \item Presentation of the main steps of proofs, possibly interleaved with text.

@@

The Gaia project~\cite{Gaia}, also maintained on \textit{Coq-Community} contains a development by José Grimm~\cite{grimm:hal-00911710}. This library -- dedicated  to the implementation in \textit{Coq (SSreflect)} of books from  Bourbaki's Elements of Mathematics --- contains a generalization of two types of ordinals (\texttt{T1} for the ordinal $\epsilon_0$ and
\texttt{T2} for $\Gamma_0$) to a larger ordinal defined by Ackermann, and relates these data structures with Bourbaki's set theory.
It is interesting to note that both libraries \textit{Gaia} and \textit{hydra-battles} contain their own (isomorphic but not identical) definitions of types \texttt{T1} and \texttt{T2}, both originated in~\cite{CantorContrib}.
Our plan is to build a  ``gate'' giving access to both the combinatorial results of \textit{hydra-battles} and the set-theoretic content of \textit{Gaia}, and make possible the transfer of theorems from one of the two libraries to the other one.

>>>>

\newcommand{\TODO}[2][]{[\textcolor{red}{TODO (#1):} \emph{#2}]}

@@

Hydra games (also known as \emph{Hydra battles}) appear in an article published in 1982 by two mathematicians, 
Laurie Kirby and Jeff Paris: \emph{Accessible Independence Results for Peano Arithmetic}~\cite{KP82}.
This article describes a game between two players: Hercules and a hydra.
A short description of the game  can be found in~\cite{bauer2008, KP82, JFLA2018paper}. One can also play with
Andrej Bauer's simulator~\cite{BauerHydra}.
In a few words:
\begin{itemize}
\item A hydra is a finite tree, traditionally presented with the root at the bottom, the leaves of which are called \emph{heads}
  (Fig~\ref{fig:round}).
\item At every round, Hercules chops off one head of the hydra. If the head is at a distance greater than 1 from the root,
  then some sub-tree $h$ of the hydra is copied a certain amount $n$ of times. The number $n$ of copies and the sub-tree $h$ may depend of the considered variant of the game
  or the time elapsed since the beginning of the fight.
  Figure~\ref{fig:round} shows an example with $n=2$.
\end{itemize}

@@

\TODO[Clément]{We haven't defined winning strategies}
\begin{theorem}
  In the Hydra game, every strategy is a winning strategy for Hercules. \label{kp:thm1}
\end{theorem}

@@

Complex formalisations and proofs are explained in an
  electronic book~\cite{HydraBook}. Whenever various reasonable choices exist, we try to present and compare the alternatives.
  For instance, Figures~\ref{fig:Ex42E0} and \ref{fig:Ex42-schutte} show two radically different proofs of the equality
  $\omega+42+\omega^2=\omega^2$. The first one is a simple proof by computation, the second one shows how this equality
  is a consequence of the axioms of the set-theoretic model  by Kurt Schütte~\cite{schutte}.

@@

 Prior stages of this project have already been presented at
 JFLA~\cite{PCiota, JFLA2018paper}.
We present recent evolutions of the library: new results, interaction within the Coq-Community project~\cite{CoqCommunity}, and documentation generated with Alectryon~\cite{alectryonpaper, alectryongithub}.

@@

\TODO[Clément]{Is there a word missing above?}
Considering measures applicable to \emph{all} battles allowed us to
focus on battles where the hydra can make an arbitrary number of copies at any time, which made our proof by contradiction artificially easier.
Unfortunately, the examples  most commonly shown in the litterature
(see for instance \cite{KP82, bauer2008, BauerHydra}) 
assume that the hydra grows $n$ copies at step $n$ of the game, which is incompatible with our proof.

@@

We now prove that Theorem~\ref{thm3} still holds with these typical battles by borrowing new combinatorial results from~\cite{KS81}.
Without loss of generality, we assume the following restrictions:

@@

We also study the number of steps of a battle:
Let $\alpha<\epsilon_0$ be an ordinal.
Using the \texttt{coq-equations} plug-in~\cite{sozeau:hal-01671777}, we define the function which returns the number of steps of the battle starting with $\alpha$ at step $i$, for  any $i$ in $\mathbb{N}$.
We prove that this number is greater or equal than
$H'_\alpha(i)-i$, where $H'$ is a slight variant of the Wainer Hardy hierarchy of rapidly growing functions~\cite{BW85, KS81, Promel2013, Wainer1970}, defined by transfinite recursion in Fig.~\ref{fig:Hprime}. \TODO[Clément]{It seems that the term ``Wainer Hardy'' hierarchy only appears in his paper and the Hydra-battles book}
From this result, we infer that the considered function is not primitive recursive in general (for $\alpha\geq\omega^\omega$).

@@

\TODO[Clément]{Highlight `Equations' and remove `omega' highlighting; change theme to something less garish}

@@

In order to prove formally that the length of the considered
kind of battles is not given by a primitive recursive function, we used the formalization of primitive recursive functions, part
of Russell O'Connor's contribution on G\"{o}del first incompleteness theorem \cite{OConnor05, Goedel}.
For this purpose, and above all by consideration of the scientific interest of this contribution, we chose to host and maintain this work within \textit{Coq-Community}.

@@

  \item Presentation of the main steps of proofs, possibly interleaved with text.
  \item \TODO[Clément]{A good part of the LaTeX support in Alectryon was developed for hydra-battles, and this was the first real application with a custom driver, too.}
  \item \TODO[Clément]{If you kept a trace of them, I would love to see a discussion of any errors that were caught by including the output systematically instead of copy-pasting it.}

@@

The Gaia project~\cite{Gaia}, also maintained on \textit{Coq-Community} contains a development by José Grimm~\cite{grimm:hal-00911710}. This library -- dedicated  to the implementation in \textit{Coq (SSreflect)} of books from  Bourbaki's Elements of Mathematics --- contains a generalization of two types of ordinals (\texttt{T1} for the ordinal $\epsilon_0$ and
\texttt{T2} for $\Gamma_0$) to a larger ordinal defined by Ackermann, and relates these data structures with Bourbaki's set theory.
It is interesting to note that both libraries \textit{Gaia} and \textit{hydra-battles} contain their own (isomorphic but not identical) definitions of types \texttt{T1} and \texttt{T2}, both originated in~\cite{CantorContrib}.
Our plan is to build a  ``gate'' \TODO[Clément]{Bridge?} giving access to both the combinatorial results of \textit{hydra-battles} and the set-theoretic content of \textit{Gaia}, and make possible the transfer of theorems from one of the two libraries to the other one.
